Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2021-07-04 source/sec_SGD.tex
Update on Overleaf.

We first show that FedAvg has an $\mathcal{O}(1/KT)$ convergence rate
for $\mu$-strongly convex and $L$-smooth objectives. The result relies on a technical improvement over the analysis in~\cite{li2019convergence}. Moreover, it
implies a distinction in communication efficiency that guarantees
this linear speedup for FedAvg with full and partial device participation.
With full participation, $E$ can be chosen as large as $\mathcal{O}(\sqrt{T/N})$
without degrading the linear speedup in the number of workers. On
the other hand, with partial participation, $E$ must be $\mathcal{O}(1)$
to guarantee $\mathcal{O}(1/KT)$ convergence.
\begin{theorem}
	\label{thm:SGD_scvx}Let $\overline{\mathbf{w}}_{T}=\sum_{k=1}^{N}p_{k}\mathbf{w}_{T}^{k}$ in FedAvg,
	$\nu_{\max}=\max_{k}Np_{k}$, and set decaying learning rates $\alpha_{t}=\frac{4}{\mu(\gamma+t)}$
	with $\gamma=\max\{32\kappa,E\}$ and $\kappa=\frac{L}{\mu}$. Then
	under Assumptions~\ref{ass:lsmooth} to \ref{ass:subgrad2} with full device participation, 
	% under Assumptions~\ref{ass:lsmooth},\ref{ass:stroncvx},\ref{ass:boundedvariance},\ref{ass:subgrad2} with full device participation, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right),
	\end{align*}
	and with partial device participation with at most $K$ sampled devices
	at each communication round, 
	\begin{align*}{\small
	\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right).}
	\end{align*}
	\label{th:scvx_sgd}
\end{theorem}
\textbf{Proof sketch.} Because our unified analyses of results in the main text follow the same framework with variations in technical details, we first give an outline of proof  for Theorem~\ref{th:scvx_sgd} to illustrate the main ideas. For full participation, the main ingredient is a recursive contraction bound 
	\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\vw^{\ast}\|^{2} &  \leq(1-\mu\alpha_{t})\mathbb{E}\|\ov{w}_{t}-\vw^{\ast}\|^{2}\\
	& +\alpha_{t}^{2}\frac{1}{N}\nu_{max}^{2}\sigma^{2}+6\alpha_{t}^{3}LE^{2}G^{2}
	\end{align*}
 where the $\mathcal{O}(\alpha_{t}^{3}E^{2}G^{2})$ term is the key improvement over the bound in~\cite{li2019convergence}, which has $\mathcal{O}(\alpha_{t}^{2}E^2 G^2)$ instead. We then use induction to obtain a non-recursive bound on $\mathbb{E}\|\ov{w}_{T}-\vw^{\ast}\|^{2}$, which is converted to a bound on $\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}$ using $L$-smoothness. For partial participation, an additional term $\mathcal{O}(\frac{1}{K} \alpha_t^2 E^2G^2)$ of leading order resulting from sampling variance is added to the contraction bound. To facilitate the understanding of our analysis, please refer to a high-level summary in Appendix~\ref{sec:app:sum}.

>>>>

We first show that FedAvg has an $\mathcal{O}(1/KT)$ convergence rate
for $\mu$-strongly convex and $L$-smooth objectives. The result relies on a technical improvement over the analysis in~\cite{li2019convergence}. Moreover, it
implies a distinction in communication efficiency that guarantees
this linear speedup for FedAvg with full and partial device participation.
With full participation, $E$ can be chosen as large as $\mathcal{O}(\sqrt{T/N})$
without degrading the linear speedup in the number of workers. On
the other hand, with partial participation, $E$ must be $\mathcal{O}(1)$
to guarantee $\mathcal{O}(1/KT)$ convergence.
\begin{theorem}
	\label{thm:SGD_scvx}Let $\overline{\mathbf{w}}_{T}=\sum_{k=1}^{N}p_{k}\mathbf{w}_{T}^{k}$ in FedAvg,
	$\nu_{\max}=\max_{k}Np_{k}$, and set decaying learning rates $\alpha_{t}=\frac{4}{\mu(\gamma+t)}$
	with $\gamma=\max\{32\kappa,E\}$ and $\kappa=\frac{L}{\mu}$. Then
	under Assumptions~\ref{ass:lsmooth} to \ref{ass:subgrad2} with full device participation, 
	% under Assumptions~\ref{ass:lsmooth},\ref{ass:stroncvx},\ref{ass:boundedvariance},\ref{ass:subgrad2} with full device participation, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right),
	\end{align*}
	and with partial device participation with at most $K$ sampled devices
	at each communication round, 
	{\small \begin{align*}
	\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa\nu_{\max}^{2}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right).
	\end{align*}}
	\label{th:scvx_sgd}
\end{theorem}
Because our unified analyses of results in the main text follow the same framework with variations in technical details, we first give an outline of proof  for Theorem~\ref{th:scvx_sgd} to illustrate the main ideas.

To facilitate the understanding of our analysis and highlight the improvement of our work comparing to prior arts, we summarize the general steps 
used in the proofs across the various settings. In this section,
we take the strongly convex case as an example to illustrate our analysis. The corresponding proof for general convex functions follows the 
same framework.

\textbf{One step progress bound} \\ 
This step establishes the progress of distance ($\|\ov{w}_{t}-\vw^{\ast}\|^{2}$) to optimal solution after one step of the FedAvg algorithm, as the following equation shows:
\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\vw^{\ast}\|^{2} & \leq \cO(\eta_t\mathbb{E}\|\ov{w}_{t}-\vw^{\ast}\|^{2} + \alpha_t^2 \sigma^2/N + \alpha_t^3E^2G^2).
\end{align*}
% \lkxcom{explain variance here and the improvement?}
The above bound consists of three main ingredients, the distance to optima
in previous step (with $\eta_t \in (0, 1)$ to obtained a contraction bound), 
the variance of stochastic gradients in local clients (second term), the variance
across different clients (third term). 
Notice that the third term in this bound is the primary source of improvement in the rate.  Comparing to the bound in~\cite{li2019convergence}, we improve 
the third term from $\cO(\alpha_t^2E^2G^2)$ to $\cO(\alpha_t^2E^2G^2)$, which enables the linear speedup in the convergence rate.

\textbf{Iterative deduction}\\
This step uses the \textit{one step progress bound} iteratively to 
connect the the current distance to optimal solution with the initial distance ($\|\ov{w}_{0}-\vw^{\ast}\|^{2}$), as follows:
\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\vw^{\ast}\|^{2} & \leq \cO(\mathbb{E}\|\ov{w}_{0}-\vw^{\ast}\|^{2} \frac{1}{T}).
\end{align*}
% \lkxcom{distinguish the difference between strongly convex and general convex?}
Then we can use the distance to optima to upper bound the optimality gap ($F(\vw_t) - F^* \leq \cO(1/T)$), as follows:
\begin{align*}
	\mathbb{E}(F(\ov{w}_{t}))-F^{\ast} & \leq\cO(\mathbb{E}\|\ov{w}_{t}-\vw^{\ast}\|^{2}). 
\end{align*}
The convergence rate of the optimality gap is equally obtained as the convergence rate of the distance to optima.

\textbf{From full participation to partial participation}\\
There are three sources of variances that affect the convergence rate.
The first two sources come from the variances of within local clients and
across clients (second and third term in one step progress bound). 
The partial participation, which involves a sampling procedure, is the third
source of variance. Therefore, comparing to the rate in full participation, this will add another term of variance into the convergence rate, where we follow a similar derivation as in~\cite{li2019convergence}.

For full participation, the main ingredient is a recursive contraction bound 
	\begin{align*}
	\mathbb{E}\|\ov{w}_{t+1}-\vw^{\ast}\|^{2} &  \leq(1-\mu\alpha_{t})\mathbb{E}\|\ov{w}_{t}-\vw^{\ast}\|^{2}\\
	& +\alpha_{t}^{2}\frac{1}{N}\nu_{max}^{2}\sigma^{2}+6\alpha_{t}^{3}LE^{2}G^{2}
	\end{align*}
 where the $\mathcal{O}(\alpha_{t}^{3}E^{2}G^{2})$ term is the key improvement over the bound in~\cite{li2019convergence}, which has $\mathcal{O}(\alpha_{t}^{2}E^2 G^2)$ instead. We then use induction to obtain a non-recursive bound on $\mathbb{E}\|\ov{w}_{T}-\vw^{\ast}\|^{2}$, which is converted to a bound on $\mathbb{E}F(\overline{\mathbf{w}}_{T})-F^{\ast}$ using $L$-smoothness. For partial participation, an additional term $\mathcal{O}(\frac{1}{K} \alpha_t^2 E^2G^2)$ of leading order resulting from sampling variance is added to the contraction bound. To facilitate the understanding of our analysis, please refer to a high-level summary in Appendix~\ref{sec:app:sum}.
